### 2024-05-23 thesis/input_files/energymarket.tex
May 23rd 2024

\section{Electricity markets}
Electricity markets have undergone rapid changes all over the world during the last three decades.
Before this revolution, the power sector was a natural monopoly. In particular, it was characterized by a high vertical integration and by very little market competition. Better technologies in both transmission, generation and distribution are the reasons for such liberal shift. The rationale is simple, a competitive market promotes efficiency and stimulates technical innovation more than a monopoly does.
\\
During the nineties various wholesale electricity markets were established; for example in Chile, Great Britain, the scandinivian block, Australia, New Zealand, the US and Canada. Consequently, new participants, each with a specific task, entered the power market scenario.
\begin{itemize}
    \item Electricity generators produce electricity to be consumed.
    \item Electricity transmission owners 
    move high voltage electricity to electric utility companies whilst ensuring safe and reliable supply.
    \item Electricity grid operators are responsible for scheduling electricity over the transmission network in order to ensure supply demand balance.
    \item Electric utilities deliver power over local lines to consumers. 
    \item Retail energy suppliers purchase electricity in the wholesale market from electricity generators and resell it to consumers.
\end{itemize}
Notice, these functions are not mutually exclusive; that is an entity can provide one or more of these services.

\subsection{The marketplace}
We can differentiate between two kinds of electricity markets: power pools and power exchanges. In power pools, generators bid the prices at which they are willing to produce at different volumes. Then, the market clearing price (MCP) is determined by intersecting the aggregated supply curve and the estimated demand, left panel figure \ref{fig:pool_vs_echange}. Power pools are created on public initiatives of governments. Conversely, power exchanges are created through private agreements between generators, distributors and traders. This is the model followed by most of european countries.
The MCP in power exchanges is determined by the intersection of the aggregate supply curve and the aggregate demand curve, right panel figure \ref{fig:pool_vs_echange}.
\\
It is worth to mention the two biggest power exchanges: Epex and NordPool. Epex operates throughout continental europe. NordPool covers the nordic and baltic countries.
\\
Moreover, we can differentiate between two popular types of auctions: uniform-price/marginal and pay as bid/discriminatory. Within the uniform price setting, suppliers offering for less than the clearing price are paid that price. Analogously, consumers bidding more than the clearing price pay that price. On the other hand, in pay as bid auctions, suppliers are paid the exact price they bid for.

@@

% outages=interruzione
\subsection{Negative prices}
It is not unusual to observe negative prices in electricity markets, even though they are rare.
The causes of negative prices are inflexible power generation plants and low demand. With inflexibility, it is mean the fact that power sources cannot be switched off and restarted quickly and efficiently.
Thus, producers are faced with the decision of either stopping and then restarting their power plant or selling their energy for a negative price; that is they pay consumers for consuming their energy.

\subsection{Nodal and zonal pricing}
Grid configuration and physical limits on electricity lines may lead to congestion. Zonal market clearing pricing (ZMCP) and local marginal price (LMP) are two types of pricing schemes utilized to handle it. The former is adopted in EU contries while the latter is used in the US, figure \ref{fig:eu_zonal} and figure \ref{fig:us_local}.
With zonal market clearing pricing, prices may differ between zones but are the same within the same area.
On the other hand, local marginal price is made up by summing the transmission congestion cost, generation marginal cost and the cost of marginal losses at different buses; buses is where a electricity line or several lines are connected.
\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/eu_zonal.png}
    \caption[Zonal pricing EU]{Zonal pricing EU \citeW{eu_zonal}}
    \label{fig:eu_zonal}
\end{figure}

>>>>

\section{Electricity markets}
Electricity markets have undergone rapid changes all over the world during the last three decades.
Before this revolution, the power sector was a natural monopoly. In particular, it was characterized by a high vertical integration and by very little market competition. Better technologies in both transmission, generation and distribution are the reasons for such liberal shift. The rationale is simple, a competitive market promotes efficiency and stimulates technical innovation more than a monopoly does.
\\
During the nineties various wholesale electricity markets were established; for example in Chile, Great Britain, the Scandinivian block, Australia, New Zealand, the US and Canada. Consequently, new participants, each with a specific task, entered the power market scenario.
\begin{itemize}
    \item Electricity generators produce electricity to be consumed.
    \item Electricity transmission owners 
    move high voltage electricity to electric utility companies whilst ensuring safe and reliable supply.
    \item Electricity grid operators are responsible for scheduling electricity over the transmission network in order to ensure supply demand balance.
    \item Electric utilities deliver power over local lines to consumers. 
    \item Retail energy suppliers purchase electricity in the wholesale market from electricity generators and resell it to consumers.
\end{itemize}
Notice, these functions are not mutually exclusive; that is an entity can provide one or more of these services.

\subsection{The marketplace}
We can differentiate between two kinds of electricity markets: power pools and power exchanges. In power pools, generators bid the prices at which they are willing to produce at different volumes. Then, the market clearing price (MCP) is determined by intersecting the aggregated supply curve and the estimated demand, left panel figure \ref{fig:pool_vs_echange}. Power pools are created on public initiatives of governments. Conversely, power exchanges are created through private agreements between generators, distributors and traders. This is the model followed by most of European countries.
The MCP in power exchanges is determined by the intersection of the aggregate supply curve and the aggregate demand curve, right panel figure \ref{fig:pool_vs_echange}.
\\
It is worth to mention the two biggest power exchanges: Epex and NordPool. Epex operates throughout continental Europe. NordPool covers the nordic and baltic countries.
\\
Moreover, we can differentiate between two popular types of auctions: uniform-price/marginal and pay-as-bid/discriminatory. Within the uniform price setting, suppliers offering for less than the clearing price are paid that price. Analogously, consumers bidding more than the clearing price pay that price. On the other hand, in pay-as-bid auctions, suppliers are paid the exact price they bid for.

@@

% outages=interruzione
\subsection{Negative prices}
It is not unusual to observe negative prices in electricity markets, even though they are rare.
The causes of negative prices are inflexible power generation plants and low demand. With inflexibility, it is meant the fact that power sources cannot be switched off and restarted quickly and efficiently.
Thus, producers are faced with the decision of either stopping and then restarting their power plant or selling their energy for a negative price; that is they pay consumers for consuming their energy.

\subsection{Nodal and zonal pricing}
Grid configuration and physical limits on electricity lines may lead to congestion. Zonal market clearing pricing (ZMCP) and local marginal price (LMP) are two types of pricing schemes utilized to handle it. The former is adopted in European contries while the latter is used in the United States, figure \ref{fig:eu_zonal} and figure \ref{fig:us_local}.
With zonal market clearing pricing, prices may differ between zones but are the same within the same area.
On the other hand, local marginal price is made up by summing the transmission congestion cost, generation marginal cost and the cost of marginal losses at different buses; buses is where a electricity line or several lines are connected.
\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/eu_zonal.png}
    \caption[Zonal pricing EU]{Zonal pricing EU \citeW{eu_zonal}}
    \label{fig:eu_zonal}
\end{figure}

### 2024-05-23 thesis/input_files/point.tex
corrected intro
May 23rd 2024

\section{Autoregressive models}
% explain their theory
% and how the procedure how they are used
% IMPORTANT: load series is not a stationary series so before applying AR we have to perform stationary tests or differencing steps
Autoregressive models are a standard approach for modelling time series data.
Before introducing this popular model in electricity point forecasting, it is important to remember that autoregressive models assume the time series to be stationary. On the other hand, load and price have been observed to be a non stationary time series. 
In a nutshell, stationarity means that the distribution of any subsequence of random variables of the stochastic process is invariant to shifts along the time dimension.
Therefore, checking stationarity is a crucial step in applying this class of models. %(unit root tests)
Furthermore, should the time series be non stationary, we can convert it to a stationary one by a combination of either differencing, detrending and/or log transforming it.
\subsection{Autoregressive integrated moving average}
ARIMA(p,d,q) models are of the form
\begin{equation}
    \hat{y}_n=\sum\limits_{i=1}^{p}\psi_i y_{n-i}^{(d)}+\sum\limits_{j=1}^{q}\theta_j \epsilon_{n-j}+\epsilon_n
\end{equation}
where the parameter p is the order of the autoregressive (AR) part, d is the degree of integrated differencing and q is the order of the moving average (MA) model. $\epsilon_{n-j}$ are the observed errors in the past while the differenced term $y_n^{(d)}$ is defined recursively as $y_n^{(d)}=y_n^{(d-1)}-y_{n-1}^{(d-1)}$.
\\
ARIMA model are compactly written as
\begin{equation}
    \psi(B)\nabla^d y_t=\theta(B)\epsilon_t
\end{equation}
where $B$ is the backward shift operator, $B^h y_t=y_{t-h}$ and $\psi(B)$ is shorthand for $\psi(B)=1-\psi_1B-\dots-\psi_p B^p$.
Similarly, $\theta(B)=1+\theta_1B+\dots+\theta_q B^q$ and $\nabla_h$ is the lag h differencing operator $\nabla_h y_t=(1-B^h)y_t=y_t-y_{t-h}$
\\
The most popular procedure for estimating this class of models is the Box-Jenkins method \cite{box2015time}.
The order p is identified by taking the lags at which the sample partial autocorrelation function (PACF) falls outside its 95\% confidence interval.
In the same way, the order q is estimated by considering the sample autocorrelation function (ACF) of the observed time series. Finally, we compare ARIMA models with different p,d,q (in a neighbourhood of the identified orders) and choose the ones minimising either AIC or BIC criteria. 
% Notice, models are obtained through the maximum likelihood estimation framework.

\subsection{Autoregressive integrated moving average with exogenous variables}
ARIMAX model adds explanatory variables to our ARIMA, hence, we have
\begin{equation}
    \hat{y}_n=\sum\limits_{k=1}^{h} \mu_i X_{n-k}+\sum\limits_{i=1}^{p}\psi_i y_{n-i}^{(d)}+\sum\limits_{j=1}^{q}\theta_j \epsilon_{n-j}+ \epsilon_{n}
\end{equation}
\subsection{Seasonal autoregressive integrated moving average and seasonal autoregressive integrated moving average with exogenous variables}
SARIMA and SARIMAX incorporate seasonalities into our time series models. The notation for these models is ARIMA$(p,d,q)(P,D,Q)_m$. The second term $(P,D,Q)_m$ represents the autoregressive structure of the seasonal pattern, where $m$ indicates the number of seasonalities of our time series.
For instance, in modelling series of hourly data with daily periodicity, we would set m=24.
Writing SARIMA in compact notation we have
\\
\begin{equation}
    \psi(B)\Psi(B^P)\nabla^d\nabla_m^Dy_t=\theta(B)\Theta(B^Q)\epsilon_t
\end{equation}
where $\Psi$ and $\Theta$ account for the $P$ and $Q$ of the seasonal pattern $m$ respectively.
\section{Generalized additive model}
% in standard regression the predicted y are normal since error are normal. But if we want to model say a y bernoulli which is the rate of success we can link the domain of bernoulli to that of gaussian with a link function
Generalized additive models (GAMs) take the form
\begin{equation}
    g[\mathbb{E}(Y_i)]=\alpha+f_1(X_1)+\dots+f_p(X_p)
\end{equation}
$f_i$ are smooth functions which are fitted using cubic smoothing splines.
g is called link function and characterizes the specific GAM model \cite{hastie2017generalized}. Approximations for $f_i$ are obtained through an iterative procedure, the backfitting algorithm.

%prophet meta
Particularly successful in time series forecasting is the Prophet model \cite{taylor2018forecasting}, developed by Meta. This additive model is capable of handling non linear trends, holiday effects and yearly weekly and daily seasonalities. Thus, the reason why we choose it as a valid benchmark to compare against.
\section{K-nearest neighbours regression}
K-nearest neighbours regression forecasts by averaging the most similar k instances in the training set \cite{macqueen1967some}. K is the algorithm hyperparameter, it stands for the number of neighbours; a small k leads to overfitting while a big k leads to underfitting. 
Since it is a metric based algorithm, it is important to normalize data in order to give equal weights to features with different scales, see \ref{appendix:normalization}.
This algorithm is made up of two decisions: first, the choice of the metric and second, the method for combining targets.

@@

Developed at the AT\&T Bell Laboratories by Vapnik et al. \cite{cortes1995support} \cite{vapnik1997support}, support vector machines (SVMs) are one of the most popular techniques within the field of statistical learning.

@@

Put differently, we would like a model to keep error less than the $\epsilon$ threshold.  

@@

    \caption{Support vector regression,\cite{learning_with_kernels}}

@@

Smola et al. \cite{smola2004tutorial} point out that considering the dual formulation makes our optimisation problem easier to solve. Doing so we have the equivalent optimisation problem

@@

Rearranging the gradient of the primal lagrangian with respect to $w$, we obtain the so called support vector expansion.

@@

Notice, the complexity of the function representation is independent of the feature space dimensionality, but depends only on the number of support vectors; that is those point $i$ for which $(\lambda_i-\lambda_i^*)\neq 0$
Moreover, we do not needd to compute $w$ explicitly in order to evaluate $f(x)$.

@@

Employing the Karush-Kuhn-Tucker conditions, $b$ can be retrieved easily. Particulalry, the condition that the product between the constraints and the dual variable has to vanish.

@@

% https://rtavenar.github.io/deep_book/book_en.pdf
Long short term memory network (LSTM) extends MLP by introducing a cell state in its layers. In this way, the current state of a cell depends on the current value $X_t$, on the previous cell activation and on the previous cell state $C_{t-1}$, see figure \ref{fig:lstm_cell} for a visual representation. 
\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/lstm_cell.png}
    \caption{Long short term memory cell}
    \label{fig:lstm_cell}
\end{figure}
The cell is responsible for deciding whether to store/forget long and short term information. To do so, each cell is made of a forget gate, an input gate and an output gate.
The forget gate takes in $h_{t-1}$ and $x_t$ and outputs a real valued vector with elements in the domain $[0,1]$, this corresponds to the ratio of information retention for each element of the cell state $C_{t-1}$.
\\
Its equation is given by
\begin{equation}
    f_t=\sigma(W_f\dot[h_{t-1},x_t]+b_f)
\end{equation}
The input gate is made up of two parts. First, a sigmoid layer decides which values of the cell state will be updated. Second, a tahn layer creates a vector of candidates $\tilde{C}_t$ to add to the current state.
\begin{equation}
    \begin{aligned}
        i_t=&\sigma(W_i\dot[h_{t-1},x_t]+b_i)
        \\
        \tilde{C}_t=&tahn(W_c\dot[h_{t-1},x_t]+b_C)
    \end{aligned}
\end{equation}
The cell state is then updated by
\begin{equation}
    C_t=f_t\odot C_{t-1}+i_t\odot \tilde{C}_t
    % odot is elementwise product
\end{equation}
Finally, we output the current cell state and the tanh activated cell state filtered by the sigmoid layer on the current input and the previous hidden state.
\begin{equation}
    \begin{aligned}
    o_t=& \sigma(W_o[h_{t-1},x_t]+b_o)
    \\
    h_t=& o_t \odot tahn(C_t)
\end{aligned}
\end{equation}
% \subsection{DeepAr} deepar é probabilistic

@@

\section{Kernel methods}
\subsection{Kernel ridge regression}
In the setting of kernel ridge regression, we aim at estimating $f$ such that $y=f(x)+\epsilon$. Given a RKHS $\mathcal{H}$, we can estimate $\hat{f}$ by solving the optmisation problem
\begin{equation}\label{eq:kernel ridge}
    \hat{f}=\argmin{f\in \mathcal{H}}\frac{1}{2}\sum\limits_{i=1}^n(y_i-f(x_i))^2+\frac{\tau}{2}\|f\|_{\mathcal{H}}^2
\end{equation}
Before proceeding, we need to introduce the representer theorem.
\begin{theorem}
    Let $k:\mathcal{X} \times \mathcal{X} \mapsto \mathbb{R}$ on a non empty set $\mathcal{X}$ with corresponding RKHS $\mathcal{H}$,
    a strictly increasing real valued function $g:[0,\infty) \to \mathbb{R}$, and the loss function $E:(\mathcal{X}\times \mathbb{R}^2)^n \to \mathbb{R} \cup \{\infty\}$.
    Consider the regularized risk functional $Risk[f]:f \mapsto  E\left((x_{1},y_{1},f(x_{1})),\ldots, (x_{n},y_{n},f(x_{n}))\right)+g\left(\| f\| \right)$. Then any minimizer of the empirical risk $Risk[f]$ can be represented as $ f^{*}(\cdot )=\sum\limits_{i=1}^{n}\alpha _{i}k(\cdot ,x_{i})$
\end{theorem}

\begin{proof}
    By orthogonal projection, any $f \in \mathcal{H}$ can be decomposed as the sum of two functions, with the first lying in the $\textrm{span}\{\phi(x_1),\dots, \phi(x_n)\}$ and the second staying in the orthogonal complement of the former. Letting $\nu$ denote a function from the orthogonal complement, we have
    \begin{equation}
        f=\sum\limits_{i=1}^{n}\alpha_i \phi(x_i)+\nu
    \end{equation}
    Applying $f$ to any point, by the reproducing property we have
    \begin{equation}
        \begin{aligned}
            f(x_j)=&\langle \phi(x_j), \sum\limits_{i=1}^{n}\alpha_i \phi(x_i)+\nu\rangle_{\mathcal{H}}
            \\
            =&\sum\limits_{i=1}^{n}\alpha_i\langle \phi(x_j), \phi(x_i)\rangle_{\mathcal{H}}
        \end{aligned}
    \end{equation}
    This means that setting $\nu=0$ does not affect the evaluation of $f$.
    Next consider the regularization term; notice, the functional $g$ is defined to be strictly monotonic in order to penalize more complex functions $f$.
    \begin{equation}
        \begin{aligned}
            g(\|f\|_{\mathcal{H}})=&g\left(\left\|\sum\limits_{i=1}^n \alpha_i \phi(x_i)+v\right\|_\mathcal{H}\right)
            \\
            =&g\left(\sqrt{\left\|\sum\limits_{i=1}^n \alpha_i \phi(x_i)+v\right\|_\mathcal{H}^2}\right)
            \\
            =&g\left(\sqrt{\left\|\sum\limits_{i=1}^n \alpha_i \phi(x_i)\right\|_{\mathcal{H}}^2+\left\|v\right\|_\mathcal{H}^2+2\left\langle\sum\limits_{i=1}^n \alpha_i \phi(x_i),\nu \right\rangle_{\mathcal{H}}}\right)
            \\
            =&g\left(\sqrt{\left\|\sum\limits_{i=1}^n \alpha_i \phi(x_i)\right\|_{\mathcal{H}}^2+\|v\|_\mathcal{H}^2}\right)
            \\
            \geq &g\left(\sqrt{\left\|\sum\limits_{i=1}^n \alpha_i \phi(x_i)\right\|_{\mathcal{H}}^2}\right)
        \end{aligned}
    \end{equation}
    Hence, we can conclude that setting $\nu=0$ decreases the second term while the first term is not affected. Thus it follows that any minimizer of the risk functional of equation \ref{eq:kernel ridge} must be of the following form
    \begin{equation}
        f^*(\cdot)=\sum\limits_{i=1}^n \alpha_i \phi(x_i)=\sum\limits_{i=1}^n \alpha_i k(\cdot, x_i)
    \end{equation}
\end{proof}

@@

\begin{figure}
    \includegraphics[width=\textwidth]{images/svr3.png}
    \caption{Gaussian RBF support vector regression}
    \label{fig:svr3}
\end{figure}
Comparing theese pictures with figure \ref{fig:svr1}, it can be concluded that introducing kernels allows support vector regression to handle the non linearities in the data.

>>>>

\section{Autoregressive models}
% explain their theory
% and how the procedure how they are used
% IMPORTANT: load series is not a stationary series so before applying AR we have to perform stationary tests or differencing steps
Autoregressive models are a standard approach for modelling time series data.
Before introducing this popular model in electricity point forecasting, it is important to remember that autoregressive models assume the time series to be stationary. On the other hand, load and price have been observed to be a non stationary time series. 
In a nutshell, stationarity means that the distribution of any subsequence of random variables of the stochastic process is invariant to shifts along the time dimension.
Therefore, checking stationarity is a crucial step in applying this class of models. %(unit root tests)
Furthermore, should the time series be non stationary, we can convert it to a stationary one by a combination of either differencing, detrending and/or log transforming it.
\subsection{Autoregressive integrated moving average}
ARIMA(p,d,q) models are of the form
\begin{equation}
    \hat{y}_n=\sum\limits_{i=1}^{p}\psi_i y_{n-i}^{(d)}+\sum\limits_{j=1}^{q}\theta_j \epsilon_{n-j}+\epsilon_n
\end{equation}
where the parameter p is the order of the autoregressive (AR) part, d is the degree of integrated differencing and q is the order of the moving average (MA) model. $\epsilon_{n-j}$ are the observed errors in the past while the differenced term $y_n^{(d)}$ is defined recursively as $y_n^{(d)}:=y_n^{(d-1)}-y_{n-1}^{(d-1)}$.
\\
ARIMA models are compactly written as
\begin{equation}
    \psi(B)\nabla^d y_t=\theta(B)\epsilon_t
\end{equation}
where $B$ is the backward shift operator, $B^h y_t=y_{t-h}$ and $\psi(B)$ is shorthand for $\psi(B)=1-\psi_1B-\dots-\psi_p B^p$.
Similarly, $\theta(B)=1+\theta_1B+\dots+\theta_q B^q$ and $\nabla_h$ is the lag $h$ differencing operator $\nabla_h y_t=(1-B^h)y_t=y_t-y_{t-h}$
\\
The most popular procedure for estimating this class of models is the Box-Jenkins method \cite{box2015time}.
The order $p$ is identified by taking the lags at which the sample partial autocorrelation function (PACF) falls outside its 95\% confidence interval.
In the same way, the order $q$ is estimated by considering the sample autocorrelation function (ACF) of the observed time series. Finally, we compare ARIMA models with different $p,d,q$ (in a neighbourhood of the identified orders) and choose the ones minimising either AIC or BIC criteria. 
% Notice, models are obtained through the maximum likelihood estimation framework.

\subsection{Autoregressive integrated moving average with exogenous variables}
ARIMAX model adds explanatory variables to our ARIMA, hence, we have
\begin{equation}
    \hat{y}_n=\sum\limits_{k=1}^{h} \mu_i X_{n-k}+\sum\limits_{i=1}^{p}\psi_i y_{n-i}^{(d)}+\sum\limits_{j=1}^{q}\theta_j \epsilon_{n-j}+ \epsilon_{n}
\end{equation}
\subsection{Seasonal autoregressive integrated moving average and seasonal autoregressive integrated moving average with exogenous variables}
SARIMA and SARIMAX incorporate seasonalities into our time series models. The notation for these models is ARIMA$(p,d,q)(P,D,Q)_m$. The second term $(P,D,Q)_m$ represents the autoregressive structure of the seasonal pattern, where $m$ indicates the number of seasonalities of our time series.
For instance, in modelling series of hourly data with daily periodicity, we would set $m=24$.
Writing SARIMA in compact notation we have
\\
\begin{equation}
    \psi(B)\Psi(B^P)\nabla^d\nabla_m^Dy_t=\theta(B)\Theta(B^Q)\epsilon_t
\end{equation}
where $\Psi$ and $\Theta$ account for the $P$ and $Q$ of the seasonal pattern $m$ respectively.
\section{Generalized additive model}
% in standard regression the predicted y are normal since error are normal. But if we want to model say a y bernoulli which is the rate of success we can link the domain of bernoulli to that of gaussian with a link function
Generalized additive models (GAMs) take the form
\begin{equation}
    g[\mathbb{E}(Y_i)]=\alpha+f_1(X_1)+\dots+f_p(X_p)
\end{equation}
$f_i$ are smooth functions which are fitted using cubic smoothing splines.
$g$ is called link function and characterizes the specific GAM model \cite{hastie2017generalized}. Approximations for $f_i$ are obtained through an iterative procedure, the backfitting algorithm.

%prophet meta
Particularly successful in time series forecasting is the Prophet model \cite{taylor2018forecasting}, developed by Meta. This additive model is capable of handling non linear trends, holiday effects and yearly weekly, and daily seasonalities. Thus, the reason why we choose it as a valid benchmark to compare against.
\section{K-nearest neighbours regression}
K-nearest neighbours regression forecasts by averaging the most similar $k$ instances in the training set \cite{macqueen1967some}. $k$ is the algorithm hyperparameter, it stands for the number of neighbours; a small $k$ leads to overfitting while a big $k$ leads to underfitting. 
Since it is a metric based algorithm, it is important to normalize data in order to give equal weights to features with different scales, see \ref{appendix:normalization}.
This algorithm is made up of two decisions: first, the choice of the metric and second, the method for combining targets.

@@

Developed at the AT\&T Bell Laboratories by Vapnik et al.\ \cite{cortes1995support} \cite{vapnik1997support}, support vector machines (SVMs) are one of the most popular techniques within the field of statistical learning.

@@

Put differently, we would like a model to keep errors less than the $\epsilon$ threshold.  

@@

    \caption{Support vector regression \cite{learning_with_kernels}}

@@

Smola et al.\ \cite{smola2004tutorial} point out that considering the dual formulation makes our optimisation problem easier to solve. Doing so we have the equivalent optimisation problem

@@

Rearranging the gradient of the primal Lagrangian with respect to $w$, we obtain the so called support vector expansion.

@@

Notice, the complexity of the function representation is independent of the feature space dimensionality, but depends only on the number of support vectors; that is those points $i$ for which $(\lambda_i-\lambda_i^*)\neq 0$
Moreover, we do not need to compute $w$ explicitly in order to evaluate $f(x)$.

@@

Employing the Karush-Kuhn-Tucker conditions, $b$ can be retrieved easily. Particularly, the condition that the product between the constraints and the dual variable has to vanish.

@@

% https://rtavenar.github.io/deep_book/book_en.pdf
Long short term memory network (LSTM) extends MLP by introducing a cell state in its layers. In this way, the current state of a cell depends on the current value $X_t$, on the previous cell activation and on the previous cell state $C_{t-1}$, see figure \ref{fig:lstm_cell} for a visual representation. 
\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/lstm_cell.png}
    \caption{Long short term memory cell}
    \label{fig:lstm_cell}
\end{figure}
The cell is responsible for deciding whether to store/forget long and short term information. To do so, each cell is made of a forget gate, an input gate and an output gate.
The forget gate takes in $h_{t-1}$ and $x_t$ and outputs a real valued vector with elements in the domain $[0,1]$, this corresponds to the ratio of information retention for each element of the cell state $C_{t-1}$.
\\
Its equation is given by
\begin{equation}
    f_t=\sigma(W_f\dot[h_{t-1},x_t]+b_f)
\end{equation}
The input gate is made up of two parts. First, a sigmoid layer decides which values of the cell state will be updated. Second, a tahn layer creates a vector of candidates $\tilde{C}_t$ to add to the current state.
\begin{equation}
    \begin{aligned}
        i_t=&\sigma(W_i\dot[h_{t-1},x_t]+b_i)
        \\
        \tilde{C}_t=&tahn(W_C\dot[h_{t-1},x_t]+b_C)
    \end{aligned}
\end{equation}
The cell state is then updated by
\begin{equation}
    C_t=f_t\odot C_{t-1}+i_t\odot \tilde{C}_t
    % odot is elementwise product
\end{equation}
Finally, we output the current cell state and the tanh activated cell state filtered by the sigmoid layer on the current input and the previous hidden state.
\begin{equation}
    \begin{aligned}
    o_t=& \sigma(W_o[h_{t-1},x_t]+b_o)
    \\
    h_t=& o_t \odot tahn(C_t)
\end{aligned}
\end{equation}
% \subsection{DeepAr} deepar é probabilistic

@@

\section{Kernel methods}
\subsection{Kernel ridge regression}
In the setting of kernel ridge regression, we aim at estimating $f$ such that $y=f(x)+\epsilon$. Given a RKHS $\mathcal{H}$, we can estimate $\hat{f}$ by solving the optmisation problem
\begin{equation}\label{eq:kernel ridge}
    \hat{f}=\argmin{f\in \mathcal{H}}\frac{1}{2}\sum\limits_{i=1}^n(y_i-f(x_i))^2+\frac{\tau}{2}\|f\|_{\mathcal{H}}^2
\end{equation}
Before proceeding, we need to introduce the representer theorem.
\begin{theorem}
    Let $k:\mathcal{X} \times \mathcal{X} \mapsto \mathbb{R}$ on a non empty set $\mathcal{X}$ with corresponding RKHS $\mathcal{H}$,
    a strictly increasing real valued function $g:[0,\infty) \to \mathbb{R}$, and the loss function $E:(\mathcal{X}\times \mathbb{R}^2)^n \to \mathbb{R} \cup \{\infty\}$.
    Consider the regularized risk functional $Risk[f]:f \mapsto  E\left((x_{1},y_{1},f(x_{1})),\ldots, (x_{n},y_{n},f(x_{n}))\right)+g\left(\| f\|_{\mathcal{H}} \right)$. Then any minimizer of the empirical risk can be represented as $ f^{*}(\cdot )=\sum\limits_{i=1}^{n}\alpha _{i}k(\cdot ,x_{i})$
\end{theorem}

\begin{proof}
    By orthogonal projection, any $f \in \mathcal{H}$ can be decomposed as the sum of two functions, with the first lying in the $\textrm{span}\{\phi(x_1),\dots, \phi(x_n)\}$ and the second staying in the orthogonal complement of the former. Letting $\nu$ denote a function from the orthogonal complement, we have
    \begin{equation}
        f=\sum\limits_{i=1}^{n}\alpha_i \phi(x_i)+\nu
    \end{equation}
    Applying $f$ to any point, by the reproducing property we have
    \begin{equation}
        \begin{aligned}
            f(x_j)=&\langle \phi(x_j), \sum\limits_{i=1}^{n}\alpha_i \phi(x_i)+\nu\rangle_{\mathcal{H}}
            \\
            =&\sum\limits_{i=1}^{n}\alpha_i\langle \phi(x_j), \phi(x_i)\rangle_{\mathcal{H}}
        \end{aligned}
    \end{equation}
    This means that setting $\nu=0$ does not affect the evaluation of $f$.
    Next consider the regularization term; notice, the functional $g$ is defined to be strictly monotonic in order to penalize more complex functions $f$.
    \begin{equation}
        \begin{aligned}
            g(\|f\|_{\mathcal{H}})=&g\left(\left\|\sum\limits_{i=1}^n \alpha_i \phi(x_i)+v\right\|_\mathcal{H}\right)
            \\
            =&g\left(\sqrt{\left\|\sum\limits_{i=1}^n \alpha_i \phi(x_i)+v\right\|_\mathcal{H}^2}\right)
            \\
            =&g\left(\sqrt{\left\|\sum\limits_{i=1}^n \alpha_i \phi(x_i)\right\|_{\mathcal{H}}^2+\left\|v\right\|_\mathcal{H}^2+2\left\langle\sum\limits_{i=1}^n \alpha_i \phi(x_i),\nu \right\rangle_{\mathcal{H}}}\right)
            \\
            =&g\left(\sqrt{\left\|\sum\limits_{i=1}^n \alpha_i \phi(x_i)\right\|_{\mathcal{H}}^2+\|v\|_\mathcal{H}^2}\right)
            \\
            \geq &g\left(\sqrt{\left\|\sum\limits_{i=1}^n \alpha_i \phi(x_i)\right\|_{\mathcal{H}}^2}\right)
        \end{aligned}
    \end{equation}
    Hence, we can conclude that setting $\nu=0$ decreases the second term while the first term is not affected. Thus, it follows that any minimizer of the risk functional of equation \ref{eq:kernel ridge} must be of the following form
    \begin{equation}
        f^*(\cdot)=\sum\limits_{i=1}^n \alpha_i \phi(x_i)=\sum\limits_{i=1}^n \alpha_i k(\cdot, x_i)
    \end{equation}
\end{proof}

@@

\begin{figure}
    \includegraphics[width=\textwidth]{images/svr3.png}
    \caption{Gaussian RBF support vector regression}
    \label{fig:svr3}
\end{figure}
Comparing these pictures with figure \ref{fig:svr1}, it can be concluded that introducing kernels allows support vector regression to handle the non linearities in the data.
